---
title: "Driving Insights: Automotive Consumer Analysis with DuckDB and Dewey Data"
subtitle: "Data Wrangling and Visualization — Course Assignment"
author: "Michael Merrill | Tristan Farrow | Henrik Harman"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    theme: flatly
    highlight-style: github
    self-contained: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
  echo: true
editor: 
  markdown: 
    wrap: 72
---


```{=html}
<!-- ============================================================
     INSTRUCTOR NOTE:
     This is a combined assignment + solution .qmd file.
     
     PART 1  (lines below this block through the horizontal rule)
     = The student-facing assignment prompt. 
     Students should complete their work in a SEPARATE .qmd file 
     modeled on this one.
     
     PART 2  (after the horizontal rule)
     = The fully worked solution walkthrough, suitable for 
     classroom review and as the Dewey Data showcase example.
     ============================================================ -->
```

# Part 1: Assignment Prompt {.tabset .tabset-fade}

------------------------------------------------------------------------

## Background

You have just joined a small automotive market research consultancy as a
junior data analyst. Your first client — a regional car dealership group
— wants to understand *who* is buying *what* kind of vehicles, and
whether demographic patterns (generation, income, gender) map onto brand
preferences and fuel-type choices, including the ongoing shift toward
electric vehicles.

Your manager hands you access to two linked datasets from **Dewey
Data**:

-   **Consumer Profiles** — one row per individual, rich demographic
    detail (generation, income tier, gender, state, homeownership,
    education, and much more)
-   **Auto Ownership** — one row per individual, up to four registered
    vehicles per person, plus brand affinity scores for major
    manufacturers

Your job is to download the data, wire it up in DuckDB, wrangle it into
an analysis-ready shape, and produce a polished multi-panel
visualization your manager can put in front of the client.

------------------------------------------------------------------------

## Setup Instructions

### Step 1 — Get Your Dewey Data Credentials

Before writing any code, you need two things from your Dewey Data
account:

1.  **Your API Key** — log in at [deweydata.io](https://deweydata.io),
    navigate to **Account → API Keys**, and copy your personal key.

2.  **The Folder IDs** for the two datasets — in the Dewey Data file
    browser, open each dataset folder and copy the folder ID from the
    URL or folder details panel. You will need:

    -   The **Consumer Profiles** folder ID
    -   The **Auto Ownership** folder ID

> ⚠️ **Never hard-code your API key directly in your .qmd file.** Store
> it as an environment variable instead (see Task 1 below). This
> protects you when pushing to GitHub.

### Step 2 — Mac-Specific Setup

Open your **Terminal** (Applications → Utilities → Terminal) and add
your API key to your shell environment file so it persists across
sessions:

``` bash
# For zsh (default on modern Macs — macOS Catalina and later)
echo 'export DEWEY_API_KEY="paste-your-key-here"' >> ~/.zshrc
source ~/.zshrc

# If you are on an older Mac using bash instead
echo 'export DEWEY_API_KEY="paste-your-key-here"' >> ~/.bash_profile
source ~/.bash_profile
```

Verify it worked:

``` bash
echo $DEWEY_API_KEY
# Should print your key, not an empty line
```

You only need to do this **once**. After that, R can read it with
`Sys.getenv("DEWEY_API_KEY")`.

### Step 2 - Windows specific

Windows stores environment variables differently than Mac. You have two
options:

**Option A (Recommended) — Set a persistent User Environment Variable
via System Settings:**

1.  Press **Windows + R**, type `sysdm.cpl`, and hit Enter to open
    System Properties.
2.  Go to the **Advanced** tab → click **Environment Variables**.
3.  Under **User variables** (top section), click **New**.
4.  Set **Variable name** to `DEWEY_API_KEY` and **Variable value** to
    your actual key.
5.  Click OK → OK → OK to save.
6.  **Restart RStudio** (or any open R sessions) so it picks up the new
    variable.

Verify it worked by running this in the R console:

``` r
Sys.getenv("DEWEY_API_KEY")
# Should print your key, not an empty string ""
```

**Option B — Use a `.Renviron` file (no system settings needed):**

1.  In RStudio, open the R console and run:

    ``` r
    usethis::edit_r_environ()
    ```

    This opens (or creates) a `.Renviron` file in your home directory.

2.  Add this line to the file:

    ```         
    DEWEY_API_KEY=paste-your-key-here
    ```

3.  Save the file and restart R (**Session → Restart R** in RStudio).

> ⚠️ **Add `.Renviron` to your `.gitignore`** — never push it to GitHub.

**Finding your file paths on Windows:**

Unlike Mac where paths look like `/Users/yourname/...`, Windows paths
look like `C:/Users/yourname/...`. In R, always use **forward slashes**
(`/`) even on Windows — backslashes (`\`) cause problems in R strings.

To find your current working directory path in R:

``` r
getwd()
# Example output: "C:/Users/yourname/Documents/my-project"
```

**If `deweyr` fails because `uvx` is not found on Windows:**

The `deweyr` package uses `uvx` internally. If you get an error about
`uvx` not being found, run this in your R console to add the typical
install location to R's PATH for the session:

``` r
# Common install location when uv is installed via the official installer
Sys.setenv(PATH = paste(
  "C:/Users/yourname/AppData/Local/uv/bin",
  Sys.getenv("PATH"),
  sep = ";"   # Windows uses semicolons to separate PATH entries
))
```

Replace `yourname` with your actual Windows username. You can find your
username by opening **Command Prompt** and typing `echo %USERNAME%`.

To make this fix permanent, add `uv`'s bin folder to your **User
Environment Variables** PATH entry using the same System Properties
dialog from Option A above.

------------------------------------------------------------------------

## Tasks

### Task 1 — Download the Data via `deweyr`

Install and load the `deweyr` package, then use it to download **both**
datasets from Dewey Data to a local folder. Store your API key and
folder IDs as environment variables or in a local `.Renviron` file —
**do not paste them directly into your code**.

Your code should:

a.  Install `deweyr` from GitHub (include this in a chunk with
    `eval: false` so it does not re-run on every render)
b.  Read your API key from the environment using `Sys.getenv()`
c.  Download the Consumer Profiles dataset to `data/consumer-profiles/`
d.  Download the Auto Ownership dataset to `data/auto-ownership/`
e.  Confirm the downloads succeeded by listing the files in each folder

**Hint:** The `dewey_download()` function accepts `api_key`,
`folder_id`, and `download_path` arguments. See the README for the full
parameter list including date-partition filtering.

------------------------------------------------------------------------

### Task 2 — Connect DuckDB and Register the Parquet Files

Once the files are downloaded, connect to an in-memory DuckDB database
and register the parquet files as virtual tables — without loading them
fully into R memory.

Your code should:

a.  Load `duckdb` and `DBI` and open an in-memory connection
b.  Use a SQL `CREATE VIEW` statement to register the Consumer Profiles
    parquet file(s) as a view called `profiles`
c.  Do the same for the Auto Ownership parquet file(s) as a view called
    `auto`
d.  Run a quick `SELECT * FROM profiles LIMIT 5` and
    `SELECT * FROM auto LIMIT 5` to confirm both views are accessible
e.  Use `duckplyr` to pull each view into R as a tibble for downstream
    wrangling

**Hint:** If there are multiple parquet files in a folder (Dewey Data
sometimes partitions large datasets), DuckDB can read them all at once
with a glob pattern: `read_parquet('data/auto-ownership/*.parquet')`

**File path note (Mac):** The path inside `read_parquet()` is relative
to your `.qmd` file's location. If you saved downloads to your Mac's
Downloads folder you can use the full path:
`read_parquet('/Users/your-mac-username/Downloads/dewey-downloads/auto-ownership/*.parquet')`
You can also find your exact path in Terminal by running `pwd` inside
the data folder.

------------------------------------------------------------------------

### Task 3 — Decode and Wrangle the Data

Both datasets use coded values for categorical variables. Before you can
analyze anything meaningfully, you need to decode the most important
ones. You will also need to reshape the Auto Ownership dataset, which
stores up to four vehicles per person in wide format (`make1`, `make2`,
`make3`, `make4`, etc.).

**3a. Decode the following columns in the Consumer Profiles dataset:**

| Column | Code → Label |
|----------------------------|--------------------------------------------|
| `generation_grpcd` | B = Baby Boomer, X = Gen X, M = Millennial, S = Silent Generation |
| `gender` | M = Male, F = Female |
| `ehi_v2` (estimated household income) | A = Under \$20K, B = \$20–\$29K, C = \$30–\$39K, D = \$40–\$49K, E = \$50–\$59K, F = \$60–\$74K, G = \$75–\$99K, H = \$100–\$124K, I = \$125–\$149K, J = \$150–\$174K, K = \$175–\$199K, L = \$200–\$249K, M = \$250K+ |
| `homeownercd` | H = Homeowner, R = Renter, U = Unknown |
| `educationcd` | A = Less than HS, B = HS Graduate, C = Some College, F = College Graduate |

**3b. Decode the following columns in the Auto Ownership dataset:**

| Column | Code → Label |
|----------------------------|--------------------------------------------|
| `fueltypecd1` | G = Gasoline, F = Flex Fuel, B = Battery Electric (EV), D = Diesel, H = Hybrid |

**3c. Pivot the Auto Ownership dataset from wide to long format** so
that each row represents one vehicle (not one person). After pivoting, a
person with three registered vehicles should appear as three rows. Your
long-format table should have at minimum: `pid`, `vehicle_slot` (1–4),
`make`, `model`, `year`, `fuel_type`, `veh_class`, `style`.

**3d. Join** the decoded long-format vehicle table to the decoded
consumer profiles table on `pid`, keeping only rows where both datasets
have a match.

------------------------------------------------------------------------

### Task 4 — Query with DuckDB SQL

Write the following queries using **raw SQL** via `DBI::dbGetQuery()`
against your DuckDB connection. Do not use dplyr for these — the point
is to practice SQL directly.

**Query A:** How many vehicles of each `veh_class` are in the dataset?
Order by count descending.

**Query B:** What is the average `in_market_new_score` by
`generation_grpcd`? Only include rows where the score is not NULL. Round
to 2 decimal places.

**Query C:** Among vehicles where `fueltypecd1` is Battery Electric
(EV), what are the top 5 most common makes? Return make and count.

**Query D:** Write a query that shows the breakdown of fuel type
(`fueltypecd1`) by generation. Your result should have one row per
generation/fuel-type combination, with a column for count and a column
for the percentage of that generation's vehicles that fuel type
represents. Round percentage to 1 decimal place.

------------------------------------------------------------------------

### Task 5 — duckplyr Wrangling

Now use `duckplyr` — which lets you write familiar `dplyr`-style code
that executes via DuckDB under the hood — to answer the following.
Reference the [duckplyr basics
guide](https://pdsr.rtemis.org/duckplyrBasics) as needed.

**5a.** Using `filter()` and `summarize()` with `.by`, find the mean
`suv_affinity` and `truck_affinity` score by generation. Arrange the
result by generation.

**5b.** Using `mutate()`, create a new column called `is_ev` that is
`TRUE` when `fueltypecd1 == "B"` (Battery Electric) and `FALSE`
otherwise. Then use `count()` grouped by `generation` and `is_ev` to
summarize EV adoption by generation.

**5c.** Using `slice_max()`, find the single individual (by `pid`) with
the highest `in_market_new_score` in each generation group. What
vehicles do they own?

------------------------------------------------------------------------

### Task 6 — Multi-Panel Visualization

Using `ggplot2`, create a **single multi-panel figure** (using
`patchwork` or `facet_wrap`) that tells a coherent story about
automotive consumer behavior from this dataset. Your figure must include
**at least three panels** and address at least two of the following
angles:

-   Vehicle class or style preferences by generation
-   EV / fuel type adoption patterns (overall or by demographic group)
-   Brand affinity scores across demographic segments
-   Income tier and vehicle class or brand relationships

Requirements:

-   Each panel must have a clear, informative title
-   Axes must be labeled (no raw column name defaults)
-   Use a consistent, clean theme across all panels
-   Include a figure caption (use Quarto's `#| fig-cap:` chunk option)
    summarizing the overall story in 1–2 sentences
-   Color palettes must be accessible (consider `scale_fill_viridis_d()`
    or `ggthemes`)

------------------------------------------------------------------------

### Task 7 — Written Summary

Write **two paragraphs** directly in your .qmd (not as a code comment)
summarizing:

1.  What you found — the key patterns in the data and what they suggest
    about the dealership client's target segments
2.  What you would want to explore next with a larger pull of this data,
    and what limitations the sample size imposes on your current
    conclusions

------------------------------------------------------------------------

### Submission

Render your `.qmd` file to HTML. Push the following files to your GitHub
repository:

-   `your-name-dewey-automotive.qmd`
-   `your-name-dewey-automotive.html`
-   `your-name-dewey-automotive.md` (if generated)

> ⚠️ **Do NOT push your `data/` folder or any parquet files to GitHub.**
> Add `data/` to your `.gitignore` file. Your API key must never appear
> in any pushed file.

Submit a link to the `.html` file on GitHub via I-Learn.

------------------------------------------------------------------------

------------------------------------------------------------------------

# Part 2: Solution Walkthrough {.tabset .tabset-fade}

> This section presents a fully worked solution to the assignment above.
> It is intended for classroom review and as a reference example for
> practitioners exploring Dewey Data with R, DuckDB, and duckplyr. Code
> is annotated throughout to explain the reasoning behind each step.

------------------------------------------------------------------------

## 0 — Package Setup

```{r}
#| label: setup
#| message: false
#| warning: false

#This code might be good, but I am running into some errors right now
#knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)

# Install deweyr from GitHub (run once — eval: false in production)
# install.packages("devtools")
# devtools::install_github("Coxabc/deweyr")

library(deweyr)       # Dewey Data download wrapper
library(duckdb)       # DuckDB database engine
library(DBI)          # Database interface (standard R DB layer)
library(duckplyr)     # dplyr-syntax interface to DuckDB
library(tidyverse)    # dplyr, tidyr, ggplot2, stringr, etc.
library(patchwork)    # Multi-panel ggplot2 composition
```

------------------------------------------------------------------------

## 1 — Downloading the Data

### 1a. API Key and Folder IDs

Your API key lives in your environment, never in your code. On a Mac,
you added it to `~/.zshrc` during setup (see Part 1 instructions). R
reads it at runtime like this:

```{r}
#| label: credentials
#| eval: false   # <-- set eval: true once your key is configured

# ── WHERE TO PUT YOUR CREDENTIALS ──────────────────────────────────────────
#
#  OPTION A (recommended): Environment variable set in ~/.zshrc (Mac)
#    Already done during setup. Just use Sys.getenv() below.
#
#  OPTION B: .Renviron file in your project root
#    Create a file called .Renviron (no extension) in your project folder.
#    Add these two lines to it — then restart R:
#
#      DEWEY_API_KEY=paste-your-api-key-here
#
#    R loads .Renviron automatically on startup.
#    ADD .Renviron TO YOUR .gitignore — never push it to GitHub.
#
# ───────────────────────────────────────────────────────────────────────────

api_key <- Sys.getenv("DEWEY_API_KEY")

# ── FOLDER IDs ──────────────────────────────────────────────────────────────
#  Find these in the Dewey Data file browser:
#  1. Log in at deweydata.io
#  2. Navigate to the dataset folder
#  3. Copy the folder ID from the URL or the folder details panel
#
#  Replace the placeholders below with your actual folder IDs:

folder_id_profiles <- "https://api.deweydata.io/api/v1/external/data/prj_ry8qacnd__cdst_gkqw4ccfiduozhia" # Consumer profile folder, this one happens to be filtered to Utah, Idaho and California
folder_id_auto     <- "https://api.deweydata.io/api/v1/external/data/prj_ry8qacnd__cdst_6vgodnvf8johkhbk" # Auto ownership folder, this one happens to be filtered down too a few common domestic and foreign brands
```

Run this in the console if it says that uvx is not found, this was
specifically for my Mac:

Sys.setenv(PATH = paste("/Users/mckennahudson/.local/bin",
Sys.getenv("PATH"), sep = ":"))

```{r}
#| output: false      # shows code but hides output
#| eval: false
#| include: false    #maybe I will take this include part out as long as it doesn't mess up the render
# Step 1: Set up your credentials
api_key            <- Sys.getenv("DEWEY_API_KEY")
folder_id_auto     <- "https://api.deweydata.io/api/v1/external/data/prj_ry8qacnd__cdst_6vgodnvf8johkhbk"      # paste your real folder ID
folder_id_profiles <- "https://api.deweydata.io/api/v1/external/data/prj_ry8qacnd__cdst_gkqw4ccfiduozhia"   # paste your real folder ID

# Step 2: Create destination folders inside your working directory
dir.create("data/auto-ownership",    recursive = TRUE, showWarnings = FALSE)
dir.create("data/consumer-profiles", recursive = TRUE, showWarnings = FALSE)

# Step 3: Download auto ownership
# Technically I don't need to run these lines anymore ebcuase now the files are safely saved on my drive
download_dewey(
  api_key       = api_key,
  folder_id     = folder_id_auto,
  download_path = "/Users/mckennahudson/DS488_DS_Consulting/data/auto-ownership"
)

# Step 4: Download consumer profiles
download_dewey(
  api_key       = api_key,
  folder_id     = folder_id_profiles,
  download_path = "/Users/mckennahudson/DS488_DS_Consulting/data/consumer-profiles"
)


#Step 5
# Check using the full absolute path instead
list.files("/Users/mckennahudson/DS488_DS_Consulting/data/auto-ownership", 
           recursive = TRUE)

list.files("/Users/mckennahudson/DS488_DS_Consulting/data/consumer-profiles", 
           recursive = TRUE)
```



------------------------------------------------------------------------

## 2 — Connect DuckDB and Register Parquet Files

After `dewey_download()` runs, your parquet files will be sitting in the
two local folders you created. DuckDB can read them directly off disk —
no need to load them into R memory first. This is one of DuckDB's
biggest advantages: you can query files that are far larger than your
available RAM.

```{r}
#| label: duckdb-connect
#| output: false      # shows code but hides output

# Open an in-memory DuckDB connection
# (DuckDB handles the file I/O itself — "in-memory" just means no separate DB file)
con <- dbConnect(duckdb())

# ── Register parquet files as DuckDB views ──────────────────────────────────
#
# The glob pattern *.parquet picks up ALL parquet files in a folder at once.
# Dewey Data sometimes splits large datasets across multiple partitioned files —
# the glob handles that automatically.
#
# IMPORTANT: these paths are relative to your .qmd file's location.
# If your .qmd is at ~/Documents/my-project/analysis.qmd, then
# "data/consumer-profiles/*.parquet" resolves to
# ~/Documents/my-project/data/consumer-profiles/*.parquet
#
# Adjust the paths below if you saved your files to a different location.
# For example, if you used an absolute path like /Users/yourname/Downloads/dewey-downloads/
# just paste that full path in place of "data/consumer-profiles".

# Register auto ownership
dbExecute(con, "
  CREATE VIEW auto AS
  SELECT * FROM read_parquet('/Users/mckennahudson/DS488_DS_Consulting/data/auto-ownership/auto-ownership-domestic-foreign-brands/*.snappy.parquet')
")

# Register consumer profiles
dbExecute(con, "
  CREATE VIEW profiles AS
  SELECT * FROM read_parquet('/Users/mckennahudson/DS488_DS_Consulting/data/consumer-profiles/idaho-utah-california-consumer-profile/*.snappy.parquet')
")

# Confirm both work
dbGetQuery(con, "SELECT * FROM auto LIMIT 3")
dbGetQuery(con, "SELECT * FROM profiles LIMIT 3")
```

Ok, now I will check that the data is good

```{r}
#| output: false      # shows code but hides output

# See exact column names as they appear in the real data (uppercase)
dbGetQuery(con, "SELECT column_name FROM information_schema.columns 
                 WHERE table_name = 'auto' LIMIT 20")

dbGetQuery(con, "SELECT column_name FROM information_schema.columns 
                 WHERE table_name = 'profiles' 
                 AND column_name IN ('PID','GENDER','GENERATION_GRPCD',
                                     'EHI_V2','STATE','AGE','HOMEOWNERCD',
                                     'EDUCATIONCD','WEALTHSCR_V2')")
```

Then join and filter within duck db before trying to pull into R

```{r}
#| output: false      # shows code but hides output



# Do the heavy join inside DuckDB — only pull key columns into R
# This avoids loading 300+ columns x 12M rows into memory
vehicles <- dbGetQuery(con, "
  SELECT 
    a.PID,
    a.MAKE1, a.MAKE2, a.MAKE3, a.MAKE4,
    a.MODEL1, a.MODEL2, a.MODEL3, a.MODEL4,
    a.YEAR1, a.YEAR2, a.YEAR3, a.YEAR4,
    a.FUELTYPECD1, a.FUELTYPECD2, a.FUELTYPECD3, a.FUELTYPECD4,
    a.VEH_CLASS1, a.VEH_CLASS2, a.VEH_CLASS3, a.VEH_CLASS4,
    a.STYLECD1, a.STYLECD2, a.STYLECD3, a.STYLECD4,
    a.SUV_AFFINITY, a.TRUCK_AFFINITY, a.SEDAN_AFFINITY,
    a.HYBRID_AFFINITY, a.TOYOTA_AFFINITY, a.FORD_AFFINITY,
    a.HONDA_AFFINITY, a.CHEVROLET_AFFINITY,
    a.IN_MARKET_NEW_SCORE, a.IN_MARKET_USED_SCORE,
    p.GENDER, p.GENERATION_GRPCD, p.EHI_V2,
    p.STATE, p.AGE, p.HOMEOWNERCD, p.EDUCATIONCD,
    p.WEALTHSCR_V2, p.MARRIEDCD
  FROM auto a
  INNER JOIN profiles p ON a.PID = p.PID
") |> as_tibble()

# How many matched?
nrow(vehicles)
n_distinct(vehicles$PID)

```

------------------------------------------------------------------------

## 3 — Decode and Wrangle the Data

### 3a–3b. Decode Categorical Codes

The datasets use compact letter codes for categorical variables. We
decode the most important ones up front so that all downstream analysis
and visualizations use human-readable labels.

```{r}
#| label: decode-profiles
#| output: false      # shows code but hides output

generation_labels <- c(
  "S" = "Silent Generation",
  "B" = "Baby Boomer", 
  "X" = "Gen X",
  "M" = "Millennial"
)

income_labels <- c(
  "A" = "Under $20K", "B" = "$20-29K", "C" = "$30-39K",
  "D" = "$40-49K",   "E" = "$50-59K", "F" = "$60-74K",
  "G" = "$75-99K",   "H" = "$100-124K","I" = "$125-149K",
  "J" = "$150-174K", "K" = "$175-199K","L" = "$200-249K",
  "M" = "$250K+"
)

fuel_labels <- c(
  "G" = "Gasoline", "F" = "Flex Fuel",
  "B" = "Battery Electric", "D" = "Diesel", "H" = "Hybrid"
)

vehicles_clean <- vehicles |>
  mutate(
    generation  = generation_labels[GENERATION_GRPCD],
    generation  = factor(generation, levels = generation_labels),
    gender      = case_when(GENDER == "M" ~ "Male",
                            GENDER == "F" ~ "Female",
                            TRUE ~ NA_character_),
    income_tier = income_labels[EHI_V2],
    income_tier = factor(income_tier, levels = income_labels, ordered = TRUE),
    fuel_type1  = fuel_labels[FUELTYPECD1],
    is_ev       = FUELTYPECD1 == "B",
    across(c(SUV_AFFINITY, TRUCK_AFFINITY, SEDAN_AFFINITY,
             HYBRID_AFFINITY, IN_MARKET_NEW_SCORE), as.numeric)
  )

glimpse(vehicles_clean)
```

Pivot to long format so that now there will only be one vehicle per row:

```{r}
#| label: Pivot long
#| output: false      # shows code but hides output

vehicles_long <- vehicles_clean |>
  pivot_longer(
    cols = matches("^(MAKE|MODEL|YEAR|FUELTYPECD|VEH_CLASS|STYLECD)[1-4]$"),
    names_to  = c(".value", "slot"),
    names_pattern = "^([A-Z_]+?)([1-4])$"
  ) |>
  rename(
    make       = MAKE,
    model      = MODEL,
    year       = YEAR,
    fuel_code  = FUELTYPECD,
    veh_class  = VEH_CLASS,
    style      = STYLECD
  ) |>
  mutate(
    fuel_type = fuel_labels[fuel_code],
    is_ev     = fuel_code == "B",
    year      = as.integer(year)
  ) |>
  filter(!is.na(make))

cat("Total vehicles (long):", nrow(vehicles_long), "\n")
cat("Unique owners:        ", n_distinct(vehicles_long$PID), "\n")
count(vehicles_long, slot)
```

So now we are building the plots

```{r}
library(patchwork)

# ── Shared theme ────────────────────────────────────────────────────────────
theme_dewey <- theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 11),
    plot.subtitle    = element_text(size = 9, color = "gray40"),
    axis.title       = element_text(size = 9),
    legend.position  = "bottom",
    legend.title     = element_text(size = 8),
    legend.text      = element_text(size = 8),
    panel.grid.minor = element_blank()
  )

gen_colors <- c(
  "Silent Generation" = "#4e79a7",
  "Baby Boomer"       = "#f28e2b",
  "Gen X"             = "#59a14f",
  "Millennial"        = "#e15759"
)

# ── Panel 1 — Top vehicle classes by generation ─────────────────────────────
p1_data <- vehicles_long |>
  filter(!is.na(generation), !is.na(veh_class)) |>
  count(generation, veh_class) |>
  group_by(generation) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  group_by(generation) |>
  slice_max(pct, n = 5) |>
  ungroup()

p1 <- ggplot(p1_data,
             aes(x = reorder(veh_class, pct),
                 y = pct,
                 fill = generation)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = gen_colors) +
  coord_flip() +
  facet_wrap(~ generation, nrow = 1) +
  labs(
    title    = "Top Vehicle Classes by Generation",
    subtitle = "Share of each generation's registered vehicles",
    x        = NULL,
    y        = "Share of vehicles"
  ) +
  theme_dewey

# ── Panel 2 — Fuel type by generation ───────────────────────────────────────
p2_data <- vehicles_long |>
  filter(!is.na(generation), !is.na(fuel_type)) |>
  count(generation, fuel_type) |>
  group_by(generation) |>
  mutate(pct = n / sum(n)) |>
  ungroup()

p2 <- ggplot(p2_data,
             aes(x = generation, y = pct, fill = fuel_type)) +
  geom_col(position = "fill", width = 0.65) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_viridis_d(option = "plasma", name = "Fuel Type", direction = -1) +
  labs(
    title    = "Fuel Type Mix by Generation",
    subtitle = "Share of each generation's vehicles by fuel type",
    x        = NULL,
    y        = "Share of vehicles"
  ) +
  theme_dewey +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

# ── Panel 3 — EV ownership by state (top 10 states) ─────────────────────────
p3_data <- vehicles_long |>
  filter(!is.na(STATE), !is.na(fuel_type)) |>
  group_by(STATE) |>
  summarize(
    pct_ev = mean(fuel_code == "B", na.rm = TRUE),
    n      = n(),
    .groups = "drop"
  ) |>
  filter(n > 1000) |>
  slice_max(pct_ev, n = 10)

p3 <- ggplot(p3_data,
             aes(x = reorder(STATE, pct_ev), y = pct_ev)) +
  geom_col(fill = "#4e79a7", width = 0.65) +
  geom_text(aes(label = scales::percent(pct_ev, accuracy = 0.1)),
            hjust = -0.1, size = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(
    title    = "EV Ownership Rate by State",
    subtitle = "States with 1,000+ vehicles in dataset",
    x        = NULL,
    y        = "Share of Battery Electric vehicles"
  ) +
  theme_dewey

# ── Combine ──────────────────────────────────────────────────────────────────
combined <- (p1 / (p2 | p3)) +
  plot_annotation(
    title   = "Who Drives What? Automotive Consumer Insights — Idaho, Utah & California",
    caption = "Source: Dewey Data — Consumer Profiles & Auto Ownership | n = 491,764 matched records",
    theme   = theme(
      plot.title   = element_text(size = 14, face = "bold"),
      plot.caption = element_text(size = 8, color = "gray50")
    )
  )

combined
```

I THOUGHT I WAS DONE HERE AT LINE 709

------------------------------------------------------------------------

## 4 — Query with DuckDB SQL

We register the wrangled data back into DuckDB and write raw SQL
queries. This mirrors what you would do in a production pipeline
querying large parquet files.

```{r}
#| label: register-wrangled

# Write the wrangled + joined tibble back into DuckDB as a table for SQL querying.
# (It came FROM DuckDB via parquet → we wrangled it in R → now we push it back
#  so we can practice raw SQL against it in Tasks 4 and 5.)
dbWriteTable(con, "vehicles", vehicles, overwrite = TRUE)
```

**Query A — Vehicle class counts:**

```{r}
#| label: sql-query-a

dbGetQuery(con, "
  SELECT VEH_CLASS1 AS vehicle_class,
         COUNT(*) AS n_vehicles
  FROM   auto
  WHERE  VEH_CLASS1 IS NOT NULL
  GROUP  BY VEH_CLASS1
  ORDER  BY n_vehicles DESC
")
```

**Query B — Average in-market new score by generation:**

```{r}
#| label: sql-query-b

dbGetQuery(con, "
  SELECT p.GENERATION_GRPCD AS generation,
         ROUND(AVG(a.IN_MARKET_NEW_SCORE), 2) AS avg_new_score,
         COUNT(*) AS n
  FROM   auto a
  JOIN   profiles p ON a.PID = p.PID
  WHERE  a.IN_MARKET_NEW_SCORE IS NOT NULL
    AND  p.GENERATION_GRPCD IS NOT NULL
  GROUP  BY p.GENERATION_GRPCD
  ORDER  BY avg_new_score DESC
")
```

**Query C — Top 5 EV makes:**

```{r}
#| label: sql-query-c

dbGetQuery(con, "
  SELECT MAKE1 AS make,
         COUNT(*) AS n_ev_vehicles
  FROM   auto
  WHERE  FUELTYPECD1 = 'B'
  GROUP  BY MAKE1
  ORDER  BY n_ev_vehicles DESC
  LIMIT  5
")
```

**Query D — Fuel type breakdown by generation:**

```{r}
#| label: sql-query-d

dbGetQuery(con, "
  WITH gen_totals AS (
    SELECT p.GENERATION_GRPCD AS generation,
           COUNT(*) AS total_in_gen
    FROM   auto a
    JOIN   profiles p ON a.PID = p.PID
    WHERE  p.GENERATION_GRPCD IS NOT NULL
    GROUP  BY p.GENERATION_GRPCD
  ),
  fuel_counts AS (
    SELECT p.GENERATION_GRPCD AS generation,
           a.FUELTYPECD1      AS fuel_type_code,
           COUNT(*) AS n
    FROM   auto a
    JOIN   profiles p ON a.PID = p.PID
    WHERE  p.GENERATION_GRPCD IS NOT NULL
      AND  a.FUELTYPECD1      IS NOT NULL
    GROUP  BY p.GENERATION_GRPCD, a.FUELTYPECD1
  )
  SELECT fc.generation,
         fc.fuel_type_code,
         fc.n,
         ROUND(100.0 * fc.n / gt.total_in_gen, 1) AS pct_of_generation
  FROM   fuel_counts fc
  JOIN   gen_totals  gt USING (generation)
  ORDER  BY fc.generation, fc.n DESC
")
```

------------------------------------------------------------------------

## 5 — duckplyr Wrangling

`duckplyr` lets us write familiar dplyr-style pipelines that execute
inside DuckDB, combining the readability of the tidyverse with the
performance of columnar query execution.

```{r}
#| label: duckplyr-setup

# Convert our joined tibble to a duckplyr data frame
vehicles_dk <- vehicles |> as_duckdb_tibble()
```

**5a — Brand affinity by generation:**

```{r}
#| label: duckplyr-5a

generation_labels <- c("S"="Silent Generation","B"="Baby Boomer",
                        "X"="Gen X","M"="Millennial")

vehicles_dk <- vehicles |> as_duckdb_tibble()

affinity_by_gen <- vehicles_dk |>
  filter(!is.na(GENERATION_GRPCD)) |>
  mutate(generation = generation_labels[GENERATION_GRPCD]) |>
  summarize(
    mean_suv_affinity   = mean(SUV_AFFINITY,  na.rm = TRUE),
    mean_truck_affinity = mean(TRUCK_AFFINITY, na.rm = TRUE),
    n = n(),
    .by = generation
  ) |>
  arrange(generation)

affinity_by_gen
```

**5b — EV adoption by generation:**

```{r}
#| label: duckplyr-5b

vehicles_long <- vehicles_long |>
  mutate(
    generation  = as.character(unname(generation)),
    income_tier = as.character(unname(income_tier)),
    fuel_type   = as.character(unname(fuel_type)),
    fuel_type1  = as.character(unname(fuel_type1))
  )

vehicles_long_dk <- vehicles_long |>
  select(-GENDER, -GENERATION_GRPCD, -EHI_V2) |>  # drop raw cols that have decoded equivalents
  as_duckdb_tibble()


ev_by_gen <- vehicles_long_dk |>
  filter(!is.na(generation)) |>
  mutate(is_ev = fuel_code == "B") |>
  count(generation, is_ev) |>
  mutate(label = if_else(is_ev, "Battery Electric", "Non-EV")) |>
  arrange(generation, is_ev)

ev_by_gen
```

**5c — Highest in-market score per generation:**

```{r}
#| label: duckplyr-5c

top_buyers <- vehicles_long_dk |>
  filter(!is.na(generation), !is.na(IN_MARKET_NEW_SCORE)) |>
  slice_max(IN_MARKET_NEW_SCORE, by = generation, n = 1, with_ties = FALSE) |>
  select(generation, PID, IN_MARKET_NEW_SCORE, make, model, year,
         fuel_type, veh_class, income_tier) |>
  arrange(generation)

top_buyers
```

------------------------------------------------------------------------

## 6 — Multi-Panel Visualization

We build a three-panel figure that tells a unified story: *How do
generation and income shape the kinds of vehicles Americans own, and
where does the EV transition stand?*

```{r}
#| label: vis-prep

# ── Shared theme ────────────────────────────────────────────────────────────
theme_dewey <- theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 11),
    plot.subtitle    = element_text(size = 9, color = "gray40"),
    axis.title       = element_text(size = 9),
    legend.position  = "bottom",
    legend.title     = element_text(size = 8),
    legend.text      = element_text(size = 8),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold", size = 9)
  )

# Generation color palette (colorblind-friendly)
gen_colors <- c(
  "Silent Generation" = "#4e79a7",
  "Baby Boomer"       = "#f28e2b",
  "Gen X"             = "#59a14f",
  "Millennial"        = "#e15759"
)
```

```{r}
#| label: panel-1
#| fig-height: 4

p1_data <- vehicles_long |>
  filter(!is.na(generation), !is.na(veh_class)) |>
  count(generation, veh_class) |>
  group_by(generation) |>
  mutate(pct = n / sum(n)) |>
  ungroup()

p1 <- ggplot(p1_data,
             aes(x = reorder(veh_class, pct, FUN = mean),  # mean across generations for stable order
                 y = pct,
                 fill = generation)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = gen_colors, name = "Generation") +
  coord_flip() +
  labs(
    title    = "Vehicle Class Preferences by Generation",
    subtitle = "Proportion of each generation's registered vehicles",
    x        = NULL,
    y        = "Share of vehicles"
  ) +
  theme_dewey
```

```{r}
#| label: panel-2
#| fig-height: 4

# Panel 2 — Fuel type by generation (stacked 100%)
p2_data <- vehicles_long |>
  filter(!is.na(generation), !is.na(fuel_type)) |>
  count(generation, fuel_type) |>
  group_by(generation) |>
  mutate(pct = n / sum(n)) |>
  ungroup()

p2 <- ggplot(p2_data,
             aes(x = generation, y = pct, fill = fuel_type)) +
  geom_col(position = "fill", width = 0.65) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_viridis_d(option = "plasma", name = "Fuel Type", direction = -1) +
  labs(
    title    = "Fuel Type Mix by Generation",
    subtitle = "100% stacked — share of each generation's vehicles by fuel type",
    x        = NULL,
    y        = "Share of vehicles"
  ) +
  theme_dewey +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r}
#| label: panel-3
#| fig-height: 4

# Panel 3 — SUV vs Truck affinity scores by income tier
# Use only the person-level data (one row per pid) to avoid double-counting
income_order <- c("Under $20K","$20-29K","$30-39K","$40-49K","$50-59K",
                  "$60-74K","$75-99K","$100-124K","$125-149K",
                  "$150-174K","$175-199K","$200-249K","$250K+")

p3_data <- vehicles_long |>
  distinct(PID, .keep_all = TRUE) |>        # one row per person to avoid double-counting
  filter(!is.na(income_tier)) |>
  mutate(income_tier = factor(income_tier, levels = income_order)) |>
  group_by(income_tier) |>
  summarize(
    mean_suv   = mean(SUV_AFFINITY,   na.rm = TRUE),
    mean_truck = mean(TRUCK_AFFINITY, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(cols = c(mean_suv, mean_truck),
               names_to  = "affinity_type",
               values_to = "mean_score") |>
  mutate(
    affinity_type = recode(affinity_type,
                           "mean_suv"   = "SUV Affinity",
                           "mean_truck" = "Truck Affinity")
  )

p3 <- ggplot(p3_data,
             aes(x = income_tier, y = mean_score,
                 color = affinity_type, group = affinity_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("SUV Affinity"   = "#4e79a7",
                                "Truck Affinity" = "#e15759"),
                     name = NULL) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(
    title    = "SUV vs. Truck Affinity Across Income Tiers",
    subtitle = "Mean affinity score (1 = low, 4 = high) by estimated household income",
    x        = "Income Tier",
    y        = "Mean Affinity Score"
  ) +
  theme_dewey
```

```{r}
#| label: combined-figure
#| fig-width: 14
#| fig-height: 12
#| fig-cap: "Automotive consumer behavior varies meaningfully across generations and income tiers: SUVs dominate across all cohorts but younger generations show stronger EV signals, while truck and SUV affinity follow distinct income curves. Data: Dewey Data Consumer Profiles + Auto Ownership datasets."

# Compose the three panels with patchwork
combined <- (p1 / p2 / p3) +
  plot_annotation(
    title   = "Who Drives What? Automotive Consumer Insights via Dewey Data",
    caption = "Source: Dewey Data — Consumer Profiles & Auto Ownership datasets\nAnalysis: Data Wrangling & Visualization Course",
    theme   = theme(
      plot.title   = element_text(size = 14, face = "bold"),
      plot.caption = element_text(size = 8, color = "gray50")
    )
  )

combined
```

------------------------------------------------------------------------

## 7 — Written Summary

### Findings

Even from this sample, several patterns are visible in the Dewey Data
automotive dataset. SUVs and trucks dominate across all generations, but
the mix shifts: Baby Boomers and Gen X skew more heavily toward
full-size trucks and larger SUVs, while Millennials show relatively
higher representation in the crossover and small-SUV categories. Battery
Electric vehicles remain a small share of the overall fleet but are
disproportionately represented among Millennials, consistent with
broader national trends in EV adoption. The brand affinity scores
reinforce this story — SUV affinity is relatively stable across income
tiers, while truck affinity peaks in middle-income households and
declines at the highest income levels, where luxury-brand SUV and sedan
affinities tend to be stronger.

### Limitations and Next Steps

The sample here contains 100 individuals per dataset, which is
sufficient for learning the tools but too small to draw confident
conclusions about population-level patterns — especially for smaller
segments like EV owners or the Silent Generation cohort. With a full
Dewey Data pull (potentially millions of records), the same DuckDB
pipeline developed here would scale without modification, since DuckDB
queries parquet files on disk rather than holding everything in R
memory. The most valuable next steps would be to (1) incorporate
geographic variables (`state`, `zip`) to explore regional variation in
EV adoption and vehicle class preference, (2) use the
`in_market_new_score` and `in_market_used_score` fields to identify
high-priority prospect segments for the dealership client, and (3) join
to external census or median-income data at the ZIP level to
cross-validate the `ehi_v2` income estimates.

------------------------------------------------------------------------

```{r}
#| label: cleanup
#| include: false

# Close the DuckDB connection when done
dbDisconnect(con, shutdown = TRUE)
```

# References {.tabset .tabset-fade}

## References

InfutorData. (2024). *Consumer Profiles* [Dataset]. Dewey Data. https://doi.org/10.82551/GVTF-Q257

InfutorData. (2024). *Auto Ownership* [Dataset]. Dewey Data. https://doi.org/10.82551/5358-6S46
